% Options for packages loaded elsewhere
% vim: wrap
% vim: linebreak
% vim: spell
\documentclass[
  12pt,
  letterpaper,
]{scrartcl}

\usepackage[
  margin=1in,
  footskip=0.5in,
  footnotesep=14pt
]{geometry}

\usepackage{setspace}
\usepackage{array}

\setlength{\parindent}{0.5in}
\setlength{\skip\footins}{10mm plus 2mm}
\setlength{\emergencystretch}{3em} % prevent overfull lines

\renewenvironment{quote}%
  {\list{}{\leftmargin=0.6in\rightmargin=0.6in}\item[]\setstretch{1.0}}%
  {\endlist\vspace{-\topsep}}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{textcomp}

\usepackage{calc} % for calculating minipage widths

\usepackage[xspace]{ellipsis}

\usepackage{soulutf8}
\setuldepth{o}
\renewcommand{\footnotesize}{\normalsize}

\addtokomafont{disposition}{\rmfamily}
\addtokomafont{section}{\normalsize\scshape}
\addtokomafont{subsection}{\normalsize}
\addtokomafont{subsubsection}{\normalsize}
\renewcommand{\thesection}{\Roman{section}} 
\renewcommand{\thesubsection}{\Alph{subsection}}
\renewcommand{\thesubsubsection}{\arabic{subsubsection}}

\renewcommand*{\sectionformat}{\thesection.\enskip}
\renewcommand*{\subsectionformat}{\makebox[0.5in]{}\thesubsection.\enskip}
\renewcommand*{\subsubsectionformat}{\makebox[1in]{}\thesubsubsection.\enskip}

\RedeclareSectionCommand[runin=false, afterindent=true,beforeskip=0.5\baselineskip,afterskip=0in]{section}
\RedeclareSectionCommand[runin=false, afterindent=true,beforeskip=0.25\baselineskip,afterskip=0in]{subsection}
\RedeclareSectionCommand[runin=false, afterindent=true,beforeskip=0.125\baselineskip,afterskip=0in]{subsubsection}

\title{Through the Legal Looking Glass: Recursion and Self-Reference in Legal Doctrine}
\author{Owen E. Healy}

\begin{document}

\begin{center}\vspace{1.0ex}{\noindent \large\textit{Through the Legal Looking Glass: Recursion and Self-Reference in Legal Doctrine}}\end{center}
  
\vspace{12pt}\noindent\begin{minipage}{4in}
  \setlength{\parskip}{6pt}
  
  \noindent By: Owen E. Healy

  \noindent Written as an application writing sample. Zero editing by any other person (or robot).
\end{minipage}

\setstretch{2.0}

\vspace{12pt}\section{Introduction}

Statements about statements (``second-order statements'') have long bothered
logicians, \textit{see} W.V.~Quine, \textit{Philosophy of Logic} 66--68 (2d
ed.~1986), so it is not surprising that they would cause trouble for judges as
well. But ``law about law'' crops up more than we might like. Take the classic
example from qualified immunity, where the court must ask not what the
law \textit{is} but what was \textit{clearly
established} at an earlier point in time. \textit{Pearson v. Callahan}, 555 U.S.~223, 231 (2009). Or, in
a proceeding collaterally attacking a conviction for ineffective assistance of
counsel, a court may need to ask whether a lawyer's assessment of the law was reasonable at the time
and whether a court would have accepted a hypothetical defense had it been raised.
\textit{Strickland v. Washington}, 466 U.S.~668, 687 (1984).
What these examples have in common is they ask the court to examine not what the law ultimately requires but the (possibly unsettled) state of lawmaking itself.

As it turns out, requiring courts to step back from their usual role of ``say[ing] what the law is,'' \textit{Marbury v. Madison}, 5 U.S.~137, 177 (1803), often causes confusion. Courts tend to conflate statements about the unsettled state of the law with statements about what the law ultimately requires. This essay explores such problems by surveying law-about-law in three discrete areas
that may, at first, seem unrelated: legal malpractice, liability insurance, and antitrust. The purpose will be to demonstrate that all three
are actually variations of a common underlying theme: that ``law about law'' is fundamentally different than ordinary legal
statements that merely assert what the law ultimately requires. Further,
these examples will show that trying to analyze second-order legal statements without
appreciating their distinct nature inevitably leads to erroneous results.


\section{Discussion}

\subsection{Legal Malpractice}

In a legal malpractice case arising out of litigation, the former client
accuses the lawyer of having litigated the case incorrectly. To prove
causation, the plaintiff must show that, under the \textit{correct} litigation strategy, the
outcome would have been more favorable. \textit{Bozelko v.~Papastavros}, 147
A.3d 1023, 1029 (Conn.~2016). The causation element is therefore law-about-law:
how would another court have behaved if presented with a different approach? \textit{See} \textit{id.}

To assist in analyzing this question, courts employ a device called the ``case within a
case'' (or ``suit within a suit''), in which the malpractice plaintiff
litigates the \textit{original} case to a new judgment, and the difference in outcomes is
taken to represent the plaintiff's damages. \textit{Witte v.~Desmarais}, 614
A.2d 116, 121 (N.H.~1992). The case-within-a-case collapses the distinction
between law \textit{about} the prior lawsuit and the law \textit{of} the prior
lawsuit. The causation shifts from what the result ``would have
been'' (law-about-law) to what it ``\textit{should have} been'' (\textit{i.e.}, what a correct interpretation of the law ultimately would have required). \textit{See}
\textit{Harline v.~Barker}, 912 P.2d 433, 440 (Utah 1996) (emphasis in
original) (quoting 2 Ronald E.~Mallen \& Jeffrey M.~Smith, \textit{Legal
Malpractice} § 27.7, at 641 (3d ed.~1989)). While the former ``law about law'' question would seem to be a more faithful embodiment of the usual but-for rule
from tort law, the latter offers a ``reflection or semblance'' of it that is easier to answer. \textit{Lieberman v.~Employers Ins.~of
Wausau}, 419 A.2d 417, 427 (N.J.~1980). Other justifications for the
case-within-a-case approach include making the prior judge's testimony
irrelevant. \textit{Phillips v.~Clancy}, 733 P.2d 300, 306
(Ariz.~Ct.~App.~1986).

But while the case-within-a-case approach is useful, it is important not to mistake it for the actual causation
element. If the original case has settled, the settlement
proceeds act as a safety net that would not be available in the counterfactual
``case within the case.'' \textit{Khalil
v.~Williams}, 53 EAL 2021 (Pa.~Aug.~3, 2021), Brief for Amici Curiae Pa.~Bar
Ass'n \textit{et al.}, at 11. But a prior settlement need not preclude the claim altogether, since it would incorrect to assume that ``damages can only be measured against the result the
client would have obtained if the case had been tried to a final judgment.''
\textit{Elizondo v.~Krist}, 415 S.W.3d 259, 263 (Tex.~2013);
\textit{Lieberman}, 419 A.2d at 427. In a malpractice case arising out of a
settlement, a court can avoid re-rolling the dice by requiring the client to
prove that ``absent malpractice, they probably would have recovered \emph{a
settlement} for more than'' the original amount. \textit{Elizondo}, 415 S.W.3d
at 270 (emphasis added). That proposition can be tested, for example, by
reference to ``settlements made under comparable circumstances.'' \textit{Id.}
Thus, while eliding the distinction between law \textit{about} the prior case
and the law \textit{of} the prior case can be useful for approximation
purposes, it is important to remember that the causation element of legal
malpractice is, ultimately, law-about-law.

Another issue that may arise is whether a judge or jury should decide causation.
If the question is how a prior \textit{jury} would have acted on different evidence,
there is general agreement that a jury should resolve that question as well.
\textit{Chocktoot v.~Smith}, 571 P.2d 1255, 1259 (Or.~1977). But when the
question is how a prior \textit{judge} would have ruled on a legal question, the
authority is reversed, and courts tend to reserve those questions for
themselves. \textit{Tinelli v.~Redl}, 199 F.3d 603, 606--07 (2d Cir.~1999);
\textit{Phillips}, 733 P.2d at 306; \textit{Thomas v.~Bethea}, 718 A.2d 1187,
1197 n.7 (Md.~1998). That result, however, is hard to square with the Seventh Amendment (and analogous state civil jury protections), and the reasoning of the cases so holding is questionable. The observation that ``a judge is best suited'' to answer the question, \textit{Royal Ins.~Co.~of Am.~v.~Miles \& Stockbridge, P.C.},~133
F.~Supp.~2d 747, 762 (D.~Md.~2001), is not usually a sufficient reason for withholding
an issue from jurors---who, after all, routinely hear difficult and complex cases.
\textit{In re U.S.~Fin.~Sec.~Litig.},~609 F.2d 411, 432 (9th Cir.~1979);
\textit{see also} \textit{Curtis v.~Loether}, 415 U.S.~189, 198 (1974) (risk of
jury prejudice ``insufficient to overcome the clear command of the Seventh
Amendment''); \textit{Granfinanciera, S.A.~v.~Nordberg}, 492 U.S.~33, 63 (1989)
(same for risk of inefficiency); \textit{Hana Financial, Inc.~v.~Hana Bank},
574 U.S.~418, 425 (2015) (same for risk that jury verdicts will be
``unpredictable''). And it is not clear that how a court \emph{would} have
acted in a counterfactual scenario is ``what the law is'' in the
\textit{Marbury} sense, \textit{see} \textit{Tinelli}, 199 F.3d at 607, as such a pronouncement is purely
retrospective and cannot govern future behavior. It may be that the tendency of
courts to reserve for themselves causation questions about judges stems
from uneasiness with the proposition that judges, being human, are subject to ordinary
cause-and-effect processes that juries can reason about.


\subsection{Liability Insurance}

A second example of law-about-law comes from liability insurance. ``The holder
of a liability insurance policy has a contractual right to payment \ldots when \ldots
the insured's liability to a third party is within the scope of the insurance
policy.'' 7A Couch on Ins.~§~103:11 (Westlaw 2022). And ``liability,'' is, of course, often a statement of legal judgment. Thus, when the insured has paid a
third-party claimant to satisfy an alleged liability, a question may
arise as to whether the liability satisfied was within the scope of the insurance
policy. \textit{See, e.g.},~\textit{Liberty Mut.~Ins.~Co.~v.~Metro.~Life
Ins.~Co.},~260 F.3d 54, 58--59 (1st Cir.~2001). For example, the insured may
pay to settle a lawsuit asserting multiple claims, some covered and some not.
\textit{Crosby Estate at Rancho Santa Fe Master Ass'n v.~Ironshore Specialty
Ins.~Co.}, 578 F.~Supp.~3d 1123, 1134--35 (S.D. Cal. 2022). In such cases, the
coverage question turns on law-about-law: whether the legal basis for liability
fits within the policy coverage. \textit{See} \textit{id.}

As an example, suppose the law in effect at the time of payment differed from that at the time of indemnity. \textit{See} \textit{Cardinal v.~State}, 107
N.E.2d 569, 577 (N.Y.~1952). Or suppose the prior law was unsettled.
\textit{See} \textit{id.} The common-law system, after all, allows for gaps in authority.
\textit{See} Helena Whalen-Bridge, \textit{The Reluctant
Comparativist:~Teaching Common Law Reasoning to Civil Law Students and the
Future of Comparative Legal Skills}, 58 J.~Legal Educ.~364, 367--68 (2008). And the coverage court may be tempted to fulfill its ordinary common
lawmaking function by filling those gaps---even though the gaps were not filled at the time of the underlying action. But, just as in qualified immunity, it should not: in resolving the coverage question, what matters is
that the law was \textit{in fact} unsettled, not how ``newly handed down decisions''
would have settled it ``long[] later.'' \textit{Cardinal}, 107 N.E.2d at
578.

Another issue that can arise is that an insured may justifiably settle a claim
even though it might have prevailed against the claimant at trial. \textit{See}
\textit{Luria Bros.~\& Co.~v.~Alliance Assurance Co.},~780 F.2d 1082, 1091 (2d
Cir.~1986). When that happens, the insurer may argue that the payment falls
outside the policy because a ``liability'' policy requires ``actual
liability.'' \textit{Catlin Specialty Ins.~Co.~v.~J.J.~White, Inc.},~387
F.~Supp.~3d 583, 591 (E.D.~Pa.~2019). By analogy to legal
malpractice, it might seem that the insured must prove the case-within-the-case
in the coverage suit. \textit{Servidone Const.~Corp.~v.~Security~Ins.~Co.~of
Hartford}, 477 N.E.2d 441, 445 (N.Y.~1985). But courts have tended to reject
that argument, and rightly so. \textit{Catlin}, 387 F.~Supp.~3d at 590 (surveying cases).
Requiring an insured who made a reasonable settlement of a disputed claim to prove
they would have lost at trial eliminates all incentive to settle, thus prolonging
litigation. \textit{Uniroyal, Inc.~v.~Home Ins.~Co.},~707 F.~Supp.~1368,
1378 (E.D.N.Y.~1988). It should be enough, rather, for the insured to show that the settlement was
reasonable and that the claim, if successful, would have resulted in covered liability. \textit{Id.}; \textit{Chicago, R.I.~\&
Pac.~Ry.~Co.~v.~United States}, 220 F.2d 939, 941 (7th Cir. 1955) (lack of
actual liability does not preclude indemnity for settlement). And as a matter of contract interpretation, an agreement to insure against ``liability'' must turn on the claims made
against the insured rather than the conduct giving rise to those claims---a ``liability'' policy is not a ``conduct'' policy.

In the malpractice context, it was noted above that courts sometimes elide the
distinction between law \textit{about} the prior litigation and the law
\textit{of} the prior litigation; doing so is administratively simple because
proving the case-within-a-case only requires the client to show what it has
maintained all along about the strength of its claim. \textit{See}
\textit{Lieberman}, 419 A.2d at 427. By contrast, asking an insured to prove a
case-within-a-case against itself would involve the insured ``turning about
face'' from contesting liability to embracing it. \textit{Uniroyal}, 707
F.~Supp.~at 1378;
\textit{see also} \textit{Lieberman}, 419 A.2d at 427 (suggesting
case-within-a-case not be used in malpractice action where there was a ``role
reversal''). Thus, recognizing that liability insurance is law-about-law is
necessary both to correctly enforce the scope of coverage and to avoid creating
a multitude of administrative problems in which settlement would be all but
impossible.

The situation becomes more confusing if coverage turns on facts neither implied
by, nor inconsistent with, actual liability.
\textit{See} \textit{Catlin}, 387 F.~Supp.~3d at 594. For example, coverage may
depend on the \textit{date} of conduct giving rise to injury, even though the third-party claimant would not have had to prove that date in order to recover. But recall that the insured may have made a reasonable settlement of a claim in which there was no injury at all---seemingly requiring the insured to
prove the date of an injury that never took taken place.
 \textit{Id.} If this sounds far-fetched, consider a particularly perplexing example: after deciding that asbestos manufacturer
W.R.~Grace \& Co.~could obtain indemnity for settlements to alleged customers
for whom it did not actually install asbestos, a coverage court observed that the
manufacturer was then seemingly required to prove ``the date of the
non-installation,'' a question about as comprehensible as the sound of one hand
clapping. \textit{Maryland Cas.~Co.~v.~W.R.~Grace \& Co.}, No.~88-cv-2613, 1996
WL 109068, at \*7 (S.D.N.Y.~Mar.~12, 1996). Courts have crafted various
solutions to this conundrum, but a general approach is not firmly established.
\textit{See} \textit{Catlin}, 387 F.~Supp.~3d at 594--96 (surveying
approaches).

Another puzzling situation occurs when the \textit{insurer} attempts to use the insured's hypothetical (rather than actual) liability to \textit{its} advantage.
In \textit{Apex Mortgage
Corp.~v.~Great Northern Insurance Co.},~972 F.3d 892 (7th Cir.~2020), an
insurer sought to disclaim coverage for an insured's settlement of a premises
liability claim based on an exclusion that would apply if the insured were ``in
possession'' of the property. \textit{Id.}~at 894. In arguing that the
insured's settlement established the applicability of the exclusion, the
insurer posited that liability under a premises theory could only attach if the
insured were in possession---and thus a claim under such a theory was
inherently outside the policy. \textit{Id.}~at 897--98. The Seventh Circuit
treated the issue as one of res judicata and determined that the insurer's
argument failed because a settlement is not a ``judicial ruling'' triggering preclusion.
\textit{Id.}~at 898. But that reasoning is doubtful. As a species of contract law, the coverage question is more accurately viewed as one of
interpretation: whether the insuring agreement was intended to cover risks related to
possession of the property. As an analogy, an insured could not avoid a fraud
exclusion in a professional negligence policy by arguing that a claim
denominated ``fraud'' was based on conduct that was \textit{in fact} only
negligent. \textit{Moscarillo v.~Professional Risk Management
Servs.,~Inc.},~921 A.2d 245, 256 (Md.~2007). Similarly, a commercial general
liability policy does not extend to claims of product defects even if the
damage were in fact inflicted by the insured's negligence. \textit{Kvaerner
Metals Div.~of Kvaerner U.S.,~Inc.~v.~Commercial Union Ins.~Co.},~908 A.2d 888,
896--97 (2006). To find coverage based on the insured's actual rather than
alleged conduct would sweep in risks the policy was never meant to cover.
\textit{See} \textit{id.}~at 899 (``To hold otherwise would be to convert a
policy for insurance into a performance bond.''). This observation is the
flip-side to the one stated above that refusing to indemnify a settlement of a
covered claim based on a lack of actual liability eliminates most of the value
of a ``liability'' policy. Although it may seem counterintuitive that reference to actual
facts (\textit{i.e.}, the truth) undermines the parties bargained-for coverage rather than clarifies it,
understanding that a liability policy is law-about-law shows this to be the
correct approach.


\subsection{Antitrust}

A final example of law-about-law comes from an area that is not usually
considered to be ``about'' legal proceedings the way malpractice and liability
insurance are: antitrust. In \textit{FTC v.~Actavis, Inc.}, 570 U.S.~136
(2013), the Supreme Court dealt with a so-called ``reverse-payment
settlement'': one in which a patentholder pays an alleged infringer not to
infringe (\textit{i.e.}, compete). In deciding whether such a settlement could give rise to a Sherman Act
violation, the \textit{Actavis} majority and dissent disagreed over whether a
competitor could violate the antitrust laws by paying to remove competition
that might have been precluded by a valid patent had the infringement suit
proceeded to judgment. \textit{See} \textit{id.}~at 164 (dissenting opinion).
In the dissent's view, if the patent were \emph{in fact} valid and infringed,
it could not be an antitrust violation to remove competition within its scope, even by direct horizontal agreement.
\textit{See} \textit{id.}~at 164, 171.

The majority correctly rejected this argument because it confuses law
\textit{about} the infringement litigation with the law \textit{of} the
infringement litigation. Within patent law, it is true that ``a patent is
either valid or invalid,'' no middle ground. \textit{Actavis}, 570 U.S.~at 171
(dissenting opinion). But when talking \textit{about} patent law, the situation is more nuanced. A patent whose validity has not been adjudicated
``may or may not be valid.'' \textit{Id.}~at 147 (majority opinion). If a cloud
of uncertainty regarding the outcome of an infringement trial might spur the
patentholder to ``permit[] the patent challenger to enter the market before the
patent expires,'' it violates the antitrust laws to pay to remove that
competition. \textit{See} \textit{id.}~at 154.

As it turns out, there is an unusually close parallel between \textit{Actavis} and the
insurer--insured scenario discussed above. In both, the actor under scrutiny
negotiates its liability to one party while simultaneously owing a duty to
protect the interest of a third party in the outcome of that negotiation.
That is, in the insurer--insured context, the insured negotiates its liability to the
claimant while owing a duty to the insurer not to pay unreasonably or in bad
faith. \textit{Uniroyal}, 707 F.~Supp.~at 1379. And in an \emph{Actavis} situation,
the alleged infringer negotiates its liability to the patentholder while owing
a duty to the consuming public not to unreasonably restrain trade. In both,
whether the negotiating party satisfies its duty is measured by its
\emph{potential} exposure rather than by litigating the case-within-a-case.
\textit{Luria}, 780 F.2d at 1091--92; \textit{Actavis}, 570 U.S.~at 157 (``[I]t
is normally not necessary to litigate patent validity to answer the antitrust
question~\ldots~.''). The following statements are therefore analogous:

{\vspace{5pt}\setstretch{1.0}%
\begin{enumerate}
  \item A liability insurer cannot avoid payment merely by showing that the
    insured would have prevailed against the underlying claimant.
    \textit{Luria}, 780 F.2d at 1091--92.
  \item An \textit{Actavis} defendant cannot avoid antitrust liability merely
    by showing that the patent would have been found valid and infringed had the 
    infringement suit proceeded to trial. \textit{See}
    \textit{Actavis}, 570 U.S.~at 147 (``[T]hat the agreement's
    `anticompetitive effects fall within the scope of the exclusionary
    potential of the patent' \ldots [does not] immunize the agreement from
    antitrust attack.'').
\end{enumerate}%
}\vspace{-12pt}%

The analogy goes further. Just as the prohibition on using actual facts to determine liability insurance
coverage cuts both ways, neither side of an \textit{Activis} dispute may measure liability using the ultimate outcome of a hypothetical infringement suit. In \textit{Actavis}, for example, the FTC did not seek to ``assess the validity of [the]
patent[] or [the] question[] of infringement by bringing [its] antitrust
suit''---that is, the FTC did not seek to collaterally attack the patent by
showing it was \textit{in fact not} valid or \textit{not} infringed. \textit{Actavis},
570 U.S.~at 164--65 (dissenting opinion). And with good reason: in defending the antitrust suit, the accused
infringer would have to ``turn about face'' from contesting infringement to
embracing it, ``markedly reduc[ing] the advantages \ldots of settling.''
\textit{Uniroyal}, 707 F.~Supp.~at 1378. Playing out the case-within-a-case of patent infringement would not
only be theoretically incorrect: it would lead to a quagmire in which
settlement would be impossible. And while it is good for competitors to respect the antitrust laws, it should not come at the cost of making patent litigation endless.

A final question is whether there is a textual justification for treating
antitrust law as law-about-law. Legal malpractice and liability insurance are
inherently ``about'' prior legal proceedings and therefore naturally law-about-law, but the
same justification does not hold for antitrust, which is about ``trade.''
\textit{See} 15 U.S.C.~§~1. The \textit{Actavis} majority took a policy-based
approach, reasoning that the Sherman Act should take account of uncertainty in
patent law because Congress would have wanted to put antitrust policy on par
with patent policy in reconciling the two statutes. 570 U.S.~at 137. As a policy matter, that may or may not make sense, but it is a stretch to say Congress adopted it through a mere prohibition on ``unreasonable'' restraints. Still, there may be a textual justification: the word ``trade'' in the Sherman
Act includes trade in property rights and therefore refers to the legal
proceedings that enforce those property rights. Thus, the Sherman Act is
textually law-about-law.


\section{Conclusion}

These are just three examples of law-about-law, but there are many more:
``clearly established'' law for purposes of 28 U.S.C. § 2254(d) or 42 U.S.C.
§ 1983, \textit{see} \textit{Pearson v.~Callahan}, 555 U.S.~223, 244 (2009),
``reasonable mistake[s] of law'' under the Fourth Amendment, \textit{Heien
v.~North Carolina}, 574 U.S.~54, 61 (2014), whether a regulation is ambiguous and
an agency's interpretation of it reasonable, \textit{Kisor v.~Wilkie}, 588 U.S.~558, 563 (2019), whether a seller misrepresented the
zoning status of a property, \textit{Lundin v.~Shimanski}, 368
N.W.2d 676, 679 (Wis.~1985), among others. Given the close parallels
illustrated between the three examples set out above---in particular the almost
perfect analogy between liability insurance and antitrust law---the same principles should govern other areas where the law talks about
law.


\end{document}

